An Evolutionary Algorithm for the Multiobjective Risk-Equity Constrained Routing Problem

### 2011-11-30 trunk/DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
added missing results of final runs (W-HypE, graph_ns3_1)

\subsection{Variation}\label{sec:variation}
As mutation operator, we suggest to shorten or lengthen the path of one or several trucks. In order to generate a new solution $s'$ from $s$, for each truck path, we draw a binary value $b\in\{0,1\}$ uniformly at random and create the new path $p'$ from the old one $p=(v_1=s^c,v_2,\ldots, v_l)$ as in \citep{horo2009a}:
\begin{itemize}
	\item if $b = 0$ and $l=:\textnormal{length}(p)\geq 2$, set $p' = (s^c, \ldots, v_{l-1})$	
	\item if $b = 1$ and $|V_{\textnormal{\scriptsize rem}}=\{v \in V \,|\, (v_{l}, v) \in A\}| \neq\emptyset$, choose $v_{l+1}$ from $V_{\textnormal{\scriptsize rem}}$ uniformly at random and set $p' = (v_0, \ldots, v_l, v_{l+1})$.
	\item otherwise, use the same path $p$ also in the new solution $s'$.
\end{itemize}

\subsection{Implementation in PISA}
The proposed algorithm has been implemented within the multiobjective optimization framework PISA \citep{bltz2003a} and is going to be made available via the official PISA web page\footnote{Until the source code is available via the PISA web page \url{http://www.tik.ee.ethz.ch/pisa/}, we made it temporarily available at \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip} as supplementary material for the review process.}. More precisely, the implementation of an algorithm/problem combination in PISA is divided into the problem-specific part (called \emph{variator module}) consisting of representation, initialization, and variation operators and the problem-independent part (called the \emph{selector module}). As selection module, any of the state-of-the-art selectors in PISA can be used off-the-shelf. In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{dapm2002a}, and W-HypE \citep{} with a focus on the first one.

After the initialization of the population with $\alpha$ individuals, those solutions are evaluated and their objective function values transferred to the selector module which then \TODO{write}

$\lambda$ individuals are chosen for variation and undergo the above described mutation operator creating the $\lambda$ so-called offspring.

 The objective vectors of the current population and these new solutions are then handed over to the selector module which decides which solutions to

\TODO{describe how PISA works}

@@

\begin{table}
\centering
\caption{\label{tab:referencepoints} Used reference points within the environmental selection of HypE}
\vspace{1em}
\begin{tabular}{lr}
	\toprule
	instance & reference point\\ \midrule
	\# commodities: 2 & $(10^8, 10^8, 10^8)$\\
	\# commodities: 3 & $(10^8, 10^8, 10^8)$\\
	\# commodities: 4 & $(10^9, 10^9, 10^9)$\\
	\bottomrule
\end{tabular}
\end{table}

\subsection{Differences Between Instances}
The basis for our experiments are three instances of the lazio network with 2 (lazio\_2), 3 (lazio\_3), and 4 commodities (lazio\_4\_1) which differ also differ in the number of trucks for the commodities. Whereas the number of trucks for lazio\_3 and lazio\_4\_1 have been chosen randomly, lazio\_2 has the same number of trucks than the lazio\_3 instance for the first two commodities.

\TODO{show results and talk about them}

\begin{figure}%
	\centering
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}%
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}%
	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
\end{figure}

\subsection{Random Initialization Versus Cost-Optimal Initialization}

\begin{figure}%
	\includegraphics[width=\columnwidth]{../experiments/randVsCost/allsolutions.eps}%
	\caption{\label{fig:allsolutions} Comparison between the random initialization (circles) and the cost-optimal one (crosses). Shown are the populations of 10 independent HypE runs after 1000 generations for all three objectives (top left) and each combination of two objectives.}
\end{figure}

\begin{figure}%
	\includegraphics[width=\columnwidth]{../experiments/randVsCost/onlynondominated.eps}%
	\caption{\label{fig:onlynondominated} Comparison between the random initialization (circles) and the cost-optimal one (crosses). Shown are only the non-dominated solutions in the final populations of 10 independent HypE runs after 1000 generations.}
\end{figure}

\subsection{Comparison With An Exact Algorithm for the Two Linear Objectives}
\TODO{add plots and describe what we compare (efficient cost/risk paths multiplied by the number of trucks); result: EA is better}

@@

\begin{figure}%
	\centering
	\includegraphics[width=0.8\columnwidth]{../experiments/randVsCost/diffPopsizes}%
	\caption{\label{fig:popsizes} Nondominated solutions of 10 independent runs after 100,000 function evaluations for HypE with a population size of 50 ($\times$), 100 ($\circ$), and 200 ($\square$) respectively.}
\end{figure}

@@

Instead of the environmental selection of HypE \citep{bz2011a}, other selection schemes can be easily used with the PISA framework and we show the resutls of two other algorithms exemplary. On the one hand, we employ the famous NSGA-II selection scheme \citep{dapm2002a} and, on the other hand, we use a generalized version of HypE towards a generalized weighted hypervolume indicator \citep{abbz2009b,bbtz2011a}. The boxplots of the resulting hypervolume indicator values after 100,000 function evaluations (population of size 100 after 1000 generations) are shown in Fig.~\ref{fig:algoComparison}. \TODO{What can be seen is that both? algorithms show a reduced performance regarding the standard hypervolume value. However, W-HypE was run in order to show how the search can be steered towards regions of the objective space that have been specified by the user. In our case, we defined the weight of the underlying weighted hypervolume indicator \citep{zbt2007a} as a box with corners XXX such that the weighted hypervolume indicator only influences BLABLABLA}

@@

>>>>

\subsection{Variation}\label{sec:variation}
As mutation operator, we suggest to shorten or lengthen the path of one or several trucks. In order to generate a new solution $s'$ from $s$, for each truck path, we draw a binary value $b\in\{0,1\}$ uniformly at random and create the new path $p'$ from the old one $p=(v_1=s^c,v_2,\ldots, v_l)$ as suggested and analyzed in \citep{horo2009a}:
\begin{itemize}
	\item if $b = 0$ and $l=:\textnormal{length}(p)\geq 2$, set $p' = (s^c, \ldots, v_{l-1})$	
	\item if $b = 1$ and $|V_{\textnormal{\scriptsize rem}}=\{v \in V \,|\, (v_{l}, v) \in A\}| \neq\emptyset$, choose $v_{l+1}$ from $V_{\textnormal{\scriptsize rem}}$ uniformly at random and set $p' = (v_0, \ldots, v_l, v_{l+1})$.
	\item otherwise, use the same path $p$ also in the new solution $s'$.
\end{itemize}

\subsection{Implementation in PISA}
The proposed algorithm has been implemented within the multiobjective optimization framework PISA \citep{bltz2003a} and is going to be made available via the official PISA web page\footnote{Until the source code is available via the PISA web page \url{http://www.tik.ee.ethz.ch/pisa/}, we made it temporarily available at \url{http://researchers.lille.inria.fr/~brockhof/downloads/hazmatvariator20111110.zip} as supplementary material for the review process.}. More precisely, the implementation of an algorithm/problem combination in PISA is divided into the problem-specific part (called \emph{variator module}) consisting of representation, initialization, and variation operators and the problem-independent part (called the \emph{selector module}). As selection module, any of the state-of-the-art selectors provided by PISA can be used off-the-shelf. In this study, we choose to use HypE \citep{bz2011a}, NSGA-II \citep{dapm2002a}, and W-HypE \citep{abbz2009b,bbtz2011a} with a focus on the first one.

After the initialization of the population with $\alpha$ individuals by the variator, their evaluation and sending the objective values to the selector module, a standard PISA run always continues in the same way. The selector decides which $\mu$ solutions to variate and the variator uses those $\mu$ solutions as starting points for the generation of $\lambda$ new search points which are again evaluated and send to the selector. In addition, the selector can decide whether solutions are stored in an archive or whether they are deleted. In the case of the selectors used in this study, no archive is maintained. The search stops after a predefined number of variator/selector iterations, also called generations, are performed. In our case, $\alpha=\mu=\lambda$, which will be also called population size hereafter. For more details about PISA, we refer to \citep{bltz2003a}.

\subsection{Hypervolume Estimation Algorithm for Multiobjective Optimization: The HypE Selector of PISA}
In order to fully describe the proposed EMO algorithm for the multiobjective risk-equity constrained routing problem, we briefly mention the working principles of the HypE selection. The main purpose of an EMO algorithm's environmental selection is to ensure convergence of the population towards the Pareto front as well as a good spread of it along the front \citep{deb2001a}. HypE persues these two goals by a two-step approach based on the hypervolume indicator \citep[see][]{ztlf2003a}.

Given the $\mu$ solutions of the old population and the $\lambda$ new ones, the best $\mu$ of them are chosen in the following way. First, a non-dominated sorting \citep{gold1989a} is performed---identifying sets of non-dominated fronts with different ranks. The solutions that are not dominated in the entire set get a rank of 1, the solutions of rank 2 are those which are non-dominated among the remaining solutions, and so forth. As long as the number of solutions is not exceeding $\mu$, entire fronts are chosen for the next iteration starting from the front with rank 1. For the first front for which the overall number of solutions exceeds $\mu$, a second criterion is used to determine the solutions to be kept for the next iteration. This decision is made according to the (expected) loss in hypervolume, which is a criterion that is refining the underlying Pareto dominance relation \citep{ztb2010a}---the main reason why many recent EMO algorithms employ the hypervolume indicator.

@@

In terms of performance assessment, we present both visual comparisons of the final populations as well as statistical results. For the statistical results, the hypervolume indicator was also used as quality measure of the final populations as it is the only known unary quality indicator  which is a refinement of the Pareto dominance relation and therefore has important theoretical properties: whenever a set of solutions dominates another set, also the hypervolume value is higher; on the contrary, if the hypervolume indicator value of a set A is worse than for set B, we know that B cannot dominate A \citep{ztlf2003a,ztb2010a}. In order to compare algorithms with respect to their hypervolume indicator values, we use both box plots\footnote{As implemented in MATLAB R2008b.} and the Wilcoxon rank sum test\footnote{As implemented in the statistical programming language R, version 2.1.3.0.}.

\subsection{Differences Between Instances}
The basis for our experiments are 5 instances of the lazio network with 2 (lazio\_2), 3 (lazio\_3), and 4 commodities (lazio\_4\_1, lazio\_4\_2, and lazio\_4\_3) which differ also in the number of trucks for the commodities. Whereas the number of trucks for lazio\_3 and lazio\_4\_1 have been chosen randomly, lazio\_2 has the same number of trucks than the lazio\_3 instance for the first two commodities. The lazio\_4\_2 instance has only 1 truck more for commodity 4 than lazio\_4\_1 and overall three times the number of trucks for each commodity than the instance lazio\_4\_3.

Figure~\ref{fig:compareAll} shows the non-dominated solutions of 10 independent HypE runs for each instance. What can be observed are three main statements. First of all, cost and risks of the found non-dominated solutions are highly correlated, as they are for each single arc in the problem formulations; therefore, the problem is not truly three-objective but rather bi-objective. Secondly, the range of the objective function values for the found non-dominated solutions follow the number of trucks in the problem instance, i.e., the more trucks a problem instance has the more costs, total and equity risk can be expected in the solutions found. However and thirdly, this relationship is not linear in the plots of Fig.~\ref{fig:compareAll} as can be seen, for example, between the lazio\_4\_2 and the lazio\_4\_3 instance. Though the number of trucks per commodity differs by a factor of 3 between them, the objective function values of the resulting non-dominated solutions after 1,000 generations are slightly further away. The reason for that is the much smaller search space for the lazio\_4\_3 instance which can be searched more exhaustively by the algorithm with the same number of function evalutions---resulting in a better convergence towards the Pareto front.

\begin{figure}
	\centering
	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/compAll_2000gens.eps}
	\caption{\label{fig:compareAll} Comparison between the different instances lazio\_2 (squares), lazio\_3 (circles), lazio\_4\_1 (diamonds), lazio\_4\_2 (crosses), and lazio\_4\_3 (stars). Shown are the non-dominated solutions in the final populations of 10 independent HypE runs with population size 50 after 2000 generations.}
\end{figure}

\subsection{Random Initialization Versus Cost-Optimal Initialization}
In order to see the impact of the two different initialization schemes \emph{random} and \emph{cost-optimal}, we performed 10 runs of HypE with both initialization schemes for 2,000 generations and a population size of 50 as above. Figure~\ref{fig:randVsCostOriginal} shows the resulting non-dominated solutions of the final populations. Surprisingly, the cost-optimal initialization could not produce other solutions than the cost-optimal one itself. \TODO{However, what can be seen is that the ...}

Boxplots can be found in Fig.~\ref{fig:costVsRandBoxPlots} ...

\begin{figure}
	\centering
	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/randVsCost_ns4_1_OriginalCost_2000gens.eps}
	\vspace{-1em}
	\caption{\label{fig:randVsCostOriginal} Comparison between the random initialization ($\circ$) and the cost-optimal one ($\times$). Shown are the non-dominated solutions in the final populations of 10 independent HypE runs with population size 50 after 2000 generations for all three objectives (top left) and each combination of two objectives on the lazio\_4\_1 instance.}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/randVsCost_ns4_1_New_2000gens.eps}
	\vspace{-1em}
	\caption{\label{fig:randVsCostNew} Comparison between the random initialization ($\circ$) and the cost-optimal one ($\times$). Shown are the non-dominated solutions in the final populations of 10 independent HypE runs with population size 50 after 2000 generations on the uncorrelated\_4 instance.}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/initBoxPlots_OriginalCosts.eps}
	\hfill
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/initBoxPlots_New.eps}
	\vspace{-1em}
	\caption{\label{fig:costVsRandBoxPlots} Comparison between the random initialization and the cost-optimal one.
	Shown are boxplots of hypervolume values for HypE with population size 50 on the lazio\_4 instance (left) and on the uncorrelated\_4 instance (right) after both 250 and 2000 generations.}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}%
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}%
	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
\end{figure}

\begin{figure}%
	\centering
	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/diffPopsizes}%
	\vspace{-1em}
	\caption{\label{fig:popsizes} Nondominated solutions of 10 independent runs after 100,000 function evaluations for HypE with a population size of 50 ($\times$), 100 ($\circ$), and 200 ($\square$) respectively.}
\end{figure}

@@

Instead of the environmental selection of HypE \citep{bz2011a}, other selection schemes can be easily used with the PISA framework and we show the resutls of two other algorithms exemplary. On the one hand, we employ the famous NSGA-II selection scheme \citep{dapm2002a} and, on the other hand, we use a generalized version of HypE towards a generalized weighted hypervolume indicator \citep{abbz2009b,bbtz2011a}. The boxplots of the resulting hypervolume indicator values after 100,000 function evaluations (population of size 100 after 1000 generations) are shown in Fig.~\ref{fig:algoComparison}. \TODO{What can be seen is that both? algorithms show a reduced performance regarding the standard hypervolume value. For NSGA-II, this is expected as its environmental selection schemes becomes less effective than in bi-objective problems when the number of objectives increases \citep{wbn2007a,bz2011a}. This is not the case for indicator-based algorithms such as HypE which were specifically proposed to circumvent the problems of algorithms optimizing non-compliant quality measures such as NSGA-II.

The reason for the reduced performance in the standard hypervolume indicator for W-Hype, however, is a different one. W-HypE was developed to steer the search towards regions of the objective space that have been specified by the user. To this end, it does not optimize the standard hypervolume indicator but a more generalized version of it, the weighted hypervolume indicator \citep{zbt2007a}. The idea behind W-HypE is that solutions in objective space regions with a higher weight will get a higher fitness due to the \emph{weighted} hypervolume they dominate. In order to be efficient also for a large number of objectives, W-HypE does not calculate the weighted hypervolume indicator values exactly (which is not possible for most of the weight functions) but instead samples them. Here, 10,000 samples are used in each iteration of the algorithm.

To show the principles of W-HypE in the application of our hazmat routing problem, we defined the weight of W-HypE's underlying weighted hypervolume indicator as constant in a box with corners $[7\cdot 10^3, 3\cdot 10^7, 5\cdot 10^5]$ and $[10^4, 4\cdot 10^7, 1.1\cdot 10^6]$ and zero otherwise. Instead of optimizing the standard hypervolume indicator as HypE, W-HypE now optimizes this weighted hypervolume indicator, resulting in better, i.e., dominating solutions in regions in or close to the box. Figure~\ref{fig:whype} shows the resulting non-dominated solutions of both algorithms with population size 25 after 10 independent runs of 4,000 generations each as well as the sampling box.}

This advantage of W-HypE over HypE within the pre-defined box can be observed both in early and late generations, cp.\ Fig.~\ref{fig:whype}.

@@

\TODO{W-HypE: number of non-dominated solutions in the sampling box: 14 for W-HypE, 6 for HypE; when overall non-dominated solutions are computed, W-HypE has 13, HypE only 1}

\begin{figure}
	\centering
	\includegraphics[width=0.75\columnwidth]{../experiments/randVsCost/WHypEVsHypE_ns4_1_OriginalCosts2.eps}
	\vspace{-1em}
	\caption{\label{fig:whype} Comparison between the standard HypE selection ($\circ$) with the one of W-HypE ($\times$) when the weighted hypervolume indicator is only sampled in the gray shaded box $[7\cdot 10^3, 10^4]\times[3\cdot 10^7, 4\cdot 10^7]\times[5\cdot 10^5, 1.1\cdot 10^6]$. Shown are all non-dominated solutions for each algorithm after 10 runs with 4,000 generations and population size 25.}
\end{figure}
